---
title: "North American Breeding Bird Survey: Data Download and Cleaning"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

## Overview

This document downloads the North American Breeding Bird Survey (BBS) dataset
from the USGS and applies expert-recommended quality filters to produce a
general-purpose cleaned dataset suitable for any downstream analysis (population
trend modeling, occupancy models, species distribution models, diversity
analyses, etc.).

The BBS is a long-running roadside point-count survey of breeding birds in the
United States and Canada, conducted annually since 1966 along thousands of
standardized routes. For full program details, see the
[USGS BBS program page](https://www.pwrc.usgs.gov/bbs/).

---

## 1. Parameters

All user-adjustable settings are defined here. Edit this block before running
the document to customize the cleaning behavior for your analysis.

```{r parameters}
# ── Year range ──────────────────────────────────────────────────────────────
# 1966 is the first BBS year for US routes; Canada began in 1967.
# 1993 is a common alternative start year because RPID=101 standardization
# and digital record-keeping improved substantially around that time.
# See: Sauer et al. (2017) https://doi.org/10.1650/CONDOR-17-83.1
YEAR_MIN <- 1993
YEAR_MAX <- as.integer(format(Sys.Date(), "%Y"))

# ── Observer filters ────────────────────────────────────────────────────────
# Exclude route-years where the observer is running that route for the first
# time. First-year observers show a consistent negative detection bias
# (see Section 6 for full rationale and citations).
EXCLUDE_FIRST_YEAR_OBSERVERS <- TRUE

# ── Nocturnal species filter ─────────────────────────────────────────────────
# BBS is a diurnal survey; owls and nightjars are poorly and inconsistently
# detected. TRUE removes them (see Section 8b).
EXCLUDE_NOCTURNAL <- TRUE

# ── Waterbird filter (optional, stricter analyses) ───────────────────────────
# BBS routes are roadside; aquatic habitats are undersampled. FALSE is
# appropriate for general-purpose datasets.
EXCLUDE_WATERBIRDS <- FALSE

# ── Shorebird filter (optional) ───────────────────────────────────────────────
EXCLUDE_SHOREBIRDS <- FALSE

# ── Output directory ─────────────────────────────────────────────────────────
OUTPUT_DIR <- "../data/bbsBayes2_clean"
```

---

## 2. Package Setup

```{r packages}
required_pkgs <- c("bbsBayes2", "dplyr", "readr", "knitr")

for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
  }
}

library(bbsBayes2)
library(dplyr)
library(readr)
library(knitr)

if (!dir.exists(OUTPUT_DIR)) {
  dir.create(OUTPUT_DIR, recursive = TRUE)
}
```

---

## 3. Download BBS Data

The [`bbsBayes2`](https://bbsbayes2.github.io/bbsBayes2/) package provides
`fetch_bbs_data()`, which downloads the official BBS dataset directly from
the [USGS ScienceBase repository](https://doi.org/10.5066/P9HE8XYJ) and
caches it locally. Subsequent calls detect the cache and skip re-downloading
unless `force = TRUE` is passed.

**Note:** The first download is approximately 500 MB and may take several
minutes depending on your connection speed.

```{r download}
if (!bbsBayes2::have_bbs_data()) {
  message("BBS data not found in cache. Downloading from USGS...")
  bbsBayes2::fetch_bbs_data()
  message("Download complete.")
} else {
  message("BBS data already cached. Skipping download.",
          " Use fetch_bbs_data(force = TRUE) to refresh.")
}
```

---

## 4. Load Data Tables

`load_bbs_data()` returns a named list with four tibbles. In bbsBayes2, the
`routes` table combines both survey-level metadata (observer, conditions,
quality flags) and route-level metadata (BCR, coordinates, route type).

| Table | Contents |
|---|---|
| `$birds` | Species count records: one row per species per route-year |
| `$routes` | Survey + route metadata: observer, conditions, quality flags, coordinates |
| `$species` | AOU code lookup with taxonomy and unidentified-taxa flags |
| `$meta` | Release and download date metadata |

```{r load-data}
bbs_raw <- bbsBayes2::load_bbs_data()

# Snapshot raw dimensions
raw_dims <- tibble::tibble(
  Table   = c("birds", "routes", "species", "meta"),
  Rows    = c(nrow(bbs_raw$birds), nrow(bbs_raw$routes),
              nrow(bbs_raw$species), nrow(bbs_raw$meta)),
  Columns = c(ncol(bbs_raw$birds), ncol(bbs_raw$routes),
              ncol(bbs_raw$species), ncol(bbs_raw$meta))
)

kable(raw_dims, caption = "Raw BBS table dimensions before any filtering")
```

---

## 5. Survey-Level Filters

All three filters below are applied to the **routes table** (which contains
survey-level quality fields). The resulting set of valid survey keys
(`state_num` + `route` + `year`) is then used in Section 7 to subset the
much larger birds table via an inner join.

We track row counts after each step for a complete audit trail.

```{r weather-init}
weather <- bbs_raw$routes
n_weather_raw <- nrow(weather)
```

### 5a. RPID = 101: Standard Protocol Only

`rpid` (Run Protocol ID) encodes the BBS protocol variant used on a given
survey. RPID 101 is the standard roadside 50-stop count. Other values
include mini-routes, off-road routes, and experimental designs. Mixing
protocols inflates variance and introduces systematic differences in
detectability that cannot be modeled away.

[Sauer et al. (2017)](https://doi.org/10.1650/CONDOR-17-83.1) restrict all
published BBS trend analyses to RPID = 101, as does every subsequent USGS
BBS analysis.

```{r filter-rpid}
weather <- weather |> filter(rpid == 101)

cat(sprintf("After rpid==101:  %d rows  (removed %d)\n",
            nrow(weather), n_weather_raw - nrow(weather)))
n_after_rpid <- nrow(weather)
```

### 5b. RunType = 1: Acceptable Survey Conditions

`run_type` is a binary quality flag assigned by USGS BBS program staff after
reviewing observer-submitted weather and protocol notes:

- **run_type = 1**: Survey met all protocol requirements (correct start time,
  acceptable wind and precipitation, complete route, etc.)
- **run_type = 0**: Survey conducted under non-standard conditions (excessive
  wind or rain, late start, incomplete route, etc.)

Including run_type = 0 surveys introduces detection heterogeneity that cannot
be adequately modeled and will bias trend estimates. All published BBS trend
analyses exclude them. See the
[USGS BBS Data Use Cautions](https://www.pwrc.usgs.gov/bbs/rawdata/PDF/WeatherCodes.pdf)
and the official data release:
[Pardieck et al., USGS ScienceBase](https://doi.org/10.5066/P9HE8XYJ).

```{r filter-runtype}
weather <- weather |> filter(run_type == 1)

cat(sprintf("After run_type==1: %d rows  (removed %d)\n",
            nrow(weather), n_after_rpid - nrow(weather)))
n_after_runtype <- nrow(weather)
```

### 5c. Year Range

Restrict to the user-specified year range. The default start year of 1993
reflects improved RPID=101 standardization and digital record-keeping quality.
Analysts requiring longer time series may set `YEAR_MIN <- 1966`, but should
note that pre-1980 data has more missing routes, higher observer turnover, and
some residual protocol differences
([Sauer et al. 2017](https://doi.org/10.1650/CONDOR-17-83.1)).

```{r filter-year}
weather <- weather |> filter(year >= YEAR_MIN, year <= YEAR_MAX)

cat(sprintf("After year [%d, %d]: %d rows  (removed %d)\n",
            YEAR_MIN, YEAR_MAX,
            nrow(weather), n_after_runtype - nrow(weather)))
n_after_year <- nrow(weather)
```

---

## 6. First-Year Observer Exclusion

Multiple studies have documented that observers running a BBS route for the
**first time** count detectably fewer birds than in subsequent years. This
startup bias likely reflects the time needed to learn a route's optimal
listening positions, local species assemblages, and difficult songs.

Key references:

- [Kendall et al. (1996)](https://doi.org/10.2307/2265700) — First documented
  the observer-learning effect quantitatively; showed removal of first-year
  observations reduced trend estimates for most species.
- [Sauer et al. (1994)](https://doi.org/10.2307/4088504) — Described
  observer-related variance in BBS data and recommended accounting for
  observer identity in modeling.
- [USGS BBS observer effects summary](https://www.pwrc.usgs.gov/bbs/) —
  Program documentation noting that first-year surveys produce lower counts.

The standard correction is to identify and remove (or flag) surveys where the
observer is running that specific route for the first time. Observer ID
(`obs_n`) is **route-specific** in USGS records — two different routes can
share an `obs_n` value for different people, so the grouping key must include
`state_num + route + obs_n`.

```{r observer-filter}
# Identify the first year each observer ran each route
observer_first_year <- weather |>
  group_by(state_num, route, obs_n) |>
  summarise(first_year = min(year), .groups = "drop")

# Join back and flag first-year surveys
weather <- weather |>
  left_join(observer_first_year, by = c("state_num", "route", "obs_n")) |>
  mutate(is_first_year_observer = (year == first_year))

n_first_yr <- sum(weather$is_first_year_observer)
cat(sprintf("First-year observer surveys identified: %d\n", n_first_yr))

if (EXCLUDE_FIRST_YEAR_OBSERVERS) {
  weather <- weather |> filter(!is_first_year_observer)
  cat(sprintf("After first-year exclusion: %d rows  (removed %d)\n",
              nrow(weather), n_after_year - nrow(weather)))
} else {
  message("First-year observers flagged but retained.",
          " Column 'is_first_year_observer' available for sensitivity analysis.")
}

n_after_observer <- nrow(weather)
```

---

## 7. Join Birds Table to Cleaned Survey Keys

Having filtered the routes table to only valid surveys, we now propagate
those filters to the birds table via an inner join on
`state_num + route + year`. Only bird records from surveys that passed all
weather/observer filters are retained.

```{r join-birds}
weather_keys <- weather |>
  select(state_num, route, year, obs_n, first_year, is_first_year_observer)

birds <- bbs_raw$birds
n_birds_raw <- nrow(birds)

birds_filtered <- birds |>
  inner_join(weather_keys, by = c("state_num", "route", "year"))

cat(sprintf("Birds before join (raw):       %d rows\n", n_birds_raw))
cat(sprintf("Birds after survey-key join:   %d rows  (removed %d)\n",
            nrow(birds_filtered), n_birds_raw - nrow(birds_filtered)))
n_birds_after_join <- nrow(birds_filtered)
```

---

## 8. Species-Level Filters

### 8a. Unidentified and Hybrid Taxa

The BBS species table includes records where identification was ambiguous —
for example, *Empidonax* sp., gull hybrids, or combined counts of
cryptic species pairs. These cannot be reliably attributed to a single
species and are excluded. USGS uses AOU codes above 48,000 for most
unidentified records; the `unid_combined` flag in the species table
provides a more explicit filter.
See the [USGS BBS species list documentation](https://www.pwrc.usgs.gov/bbs/rawdata/).

```{r filter-unid}
species_tbl <- bbs_raw$species

# Attempt to use the unid_combined flag; fall back to AOU < 48000 if absent
if ("unid_combined" %in% names(species_tbl)) {
  valid_aou <- species_tbl |>
    filter(unid_combined == 0) |>
    pull(aou)
} else {
  message("Column 'unid_combined' not found; using AOU < 48000 as fallback.")
  valid_aou <- species_tbl |>
    filter(aou < 48000) |>
    pull(aou)
}

n_before_unid <- nrow(birds_filtered)
birds_filtered <- birds_filtered |> filter(aou %in% valid_aou)

cat(sprintf("After unidentified taxa removal: %d rows  (removed %d)\n",
            nrow(birds_filtered), n_before_unid - nrow(birds_filtered)))
```

### 8b. Nocturnal Species (Owls and Nightjars)

The BBS protocol begins approximately 30 minutes before local sunrise and
covers 50 roadside stops over roughly 4.5 hours — it is fundamentally a
**diurnal** survey. Owls (Strigiformes) and nightjars (Caprimulgiformes) may
be heard incidentally at the earliest stops but are not systematically
detected, and their apparent BBS trends more likely reflect variation in
ambient light conditions and observer behavior than true population change.

[Robbins et al. (1986)](https://doi.org/10.2307/3830552) — the foundational
BBS methods paper — explicitly listed these groups as poorly sampled. This
exclusion is standard practice in all published BBS trend analyses; see also
the [bbsBayes2 species list rationale](https://bbsbayes2.github.io/bbsBayes2/).

AOU ranges used:

| Group | AOU range | Example species |
|---|---|---|
| Barn Owl (*Tytonidae*) | 3600 | Barn Owl |
| Owls (*Strigidae*) | 3650–3810 | Great Horned, Barred, Saw-whet |
| Nightjars (*Caprimulgidae*) | 4160–4210 | Whip-poor-will, Common Nighthawk |

```{r filter-nocturnal}
if (EXCLUDE_NOCTURNAL) {
  # Verify which species fall in these ranges against the actual species table
  nocturnal_check <- species_tbl |>
    filter((aou >= 3600 & aou <= 3810) | (aou >= 4160 & aou <= 4210)) |>
    select(aou, english) |>
    arrange(aou)

  cat("Species removed as nocturnal:\n")
  print(nocturnal_check, n = Inf)

  n_before_noct <- nrow(birds_filtered)
  birds_filtered <- birds_filtered |>
    filter(
      !(aou >= 3600 & aou <= 3810),   # Barn Owl + Strigidae owls
      !(aou >= 4160 & aou <= 4210)    # Caprimulgidae nightjars
    )

  cat(sprintf("\nAfter nocturnal species removal: %d rows  (removed %d)\n",
              nrow(birds_filtered), n_before_noct - nrow(birds_filtered)))
}
```

### 8c. Waterbirds (Optional)

BBS roadside routes systematically undersample aquatic habitats. Loons,
grebes, pelicans, cormorants, herons, and related waterbirds (AOU < 200,
approximately) are detected far less reliably than on dedicated waterbird
surveys. [Link & Sauer (2002)](https://doi.org/10.1890/0012-9658(2002)083[2832:AHMOPCA]2.0.CO;2)
discuss the detectability assumptions underlying BBS analyses and note that
habitat-associated detectability heterogeneity is a key concern.

This filter is **disabled by default** (`EXCLUDE_WATERBIRDS <- FALSE`) because
many analysts retain waterbirds for general richness or community composition
work. Enable it for analyses where detectability or habitat association is
central.

```{r filter-waterbirds}
if (EXCLUDE_WATERBIRDS) {
  n_before_water <- nrow(birds_filtered)
  birds_filtered <- birds_filtered |> filter(aou >= 200)
  cat(sprintf("After waterbird removal (AOU < 200): %d rows  (removed %d)\n",
              nrow(birds_filtered), n_before_water - nrow(birds_filtered)))
  message("NOTE: AOU < 200 cutoff is approximate. Verify against species list.")
}
```

### 8d. Shorebirds (Optional)

Shorebirds (Charadriiformes, approximately AOU 2390–2980) are also
undersampled by BBS methods for the same habitat-detectability reasons as
waterbirds. This filter is disabled by default.

```{r filter-shorebirds}
if (EXCLUDE_SHOREBIRDS) {
  n_before_shore <- nrow(birds_filtered)
  birds_filtered <- birds_filtered |>
    filter(!(aou >= 2390 & aou <= 2980))
  cat(sprintf("After shorebird removal (AOU 2390–2980): %d rows  (removed %d)\n",
              nrow(birds_filtered), n_before_shore - nrow(birds_filtered)))
}
```

---

## 9. Attach Route Metadata

Join route-level covariates (coordinates, route type, active status) to the
cleaned bird records. In bbsBayes2, this information is in `bbs_raw$routes`;
we deduplicate to one row per route before joining.

```{r attach-routes}
route_meta <- bbs_raw$routes |>
  select(state_num, route, latitude, longitude, route_type_id, active) |>
  distinct()

bbs_clean <- birds_filtered |>
  left_join(route_meta, by = c("state_num", "route"))

n_missing_route <- sum(is.na(bbs_clean$latitude))
if (n_missing_route > 0) {
  warning(sprintf("%d records could not be matched to route metadata.",
                  n_missing_route))
}
```

---

## 10. Summary Statistics

```{r summary}
summary_stats <- tibble::tibble(
  Metric  = c("Total bird count records",
               "Unique species (AOU codes)",
               "Unique routes",
               "Year range",
               "BCRs represented"),
  Value   = c(
    format(nrow(bbs_clean), big.mark = ","),
    format(n_distinct(bbs_clean$aou), big.mark = ","),
    format(n_distinct(paste(bbs_clean$state_num, bbs_clean$route)), big.mark = ","),
    paste(min(bbs_clean$year), "–", max(bbs_clean$year)),
    as.character(n_distinct(bbs_clean$bcr))
  )
)

kable(summary_stats, caption = "Cleaned dataset summary")

# Filter audit table
audit <- tibble::tibble(
  Step = c(
    "Raw birds table",
    "After survey-key join (rpid, run_type, year, observer filters)",
    "After species-level filters",
    "Final cleaned dataset"
  ),
  Rows = c(
    n_birds_raw,
    n_birds_after_join,
    nrow(birds_filtered),
    nrow(bbs_clean)
  )
) |>
  mutate(
    `Rows removed` = c(0, -diff(Rows)),
    `% of raw` = round(Rows / n_birds_raw * 100, 1)
  )

kable(audit, caption = "Filter audit: rows retained at each step",
      format.args = list(big.mark = ","))
```

---

## 11. Save Outputs

```{r save}
# Primary analysis file (preserves all R data types)
rds_path <- file.path(OUTPUT_DIR, "bbs_birds_clean.rds")
saveRDS(bbs_clean, file = rds_path)
message("Saved: ", rds_path)

# Compressed CSV for interoperability (Python, Stan, Excel, etc.)
csv_path <- file.path(OUTPUT_DIR, "bbs_birds_clean.csv.gz")
write_csv(bbs_clean, file = csv_path)
message("Saved: ", csv_path, "  (gzip compressed)")

# Cleaned weather/survey metadata for observer-level modeling
weather_path <- file.path(OUTPUT_DIR, "bbs_weather_clean.rds")
saveRDS(weather, file = weather_path)
message("Saved: ", weather_path)

# Reproducibility log
params_log <- tibble::tibble(
  parameter = c("YEAR_MIN", "YEAR_MAX",
                "EXCLUDE_FIRST_YEAR_OBSERVERS",
                "EXCLUDE_NOCTURNAL",
                "EXCLUDE_WATERBIRDS",
                "EXCLUDE_SHOREBIRDS",
                "date_run",
                "R_version",
                "bbsBayes2_version"),
  value = c(YEAR_MIN, YEAR_MAX,
            EXCLUDE_FIRST_YEAR_OBSERVERS,
            EXCLUDE_NOCTURNAL,
            EXCLUDE_WATERBIRDS,
            EXCLUDE_SHOREBIRDS,
            as.character(Sys.Date()),
            as.character(getRversion()),
            as.character(packageVersion("bbsBayes2")))
)

params_path <- file.path(OUTPUT_DIR, "run_parameters.csv")
write_csv(params_log, file = params_path)
message("Saved: ", params_path)

message("\nDone. Load cleaned data with: bbs <- readRDS('", rds_path, "')")
```

---

## Notes for Downstream Analysis

**Zero-filling**: This output is a *detections-only* table — unobserved
species on a route-year are implicit zeros. Most population trend and
occupancy models require explicit zero rows. Zero-filling is deliberately
omitted here because (1) it inflates file size 10–50× and (2) the correct
approach depends on your target species list and spatial extent. When ready:

```r
bbs_complete <- tidyr::complete(
  bbs_clean,
  tidyr::nesting(state_num, route, year, obs_n),
  aou,
  fill = list(stop_total = 0)
)
```

**Column name compatibility**: bbsBayes2 returns snake_case column names
(`aou`, `state_num`). If downstream joins fail, standardize with
`names(df) <- tolower(names(df))`.

**Further stratification**: For trend modeling, the cleaned data can be
passed to `bbsBayes2::stratify(by = "bcr")` after subsetting to a single
species, or used directly with `lme4`, `Stan`, or `JAGS` models.
